%
% Copyright © 2018 Peeter Joot.  All Rights Reserved.
% Licenced as described in the file LICENSE under the root directory of this GIT repository.
%
%{
\input{../latex/blogpost.tex}
\renewcommand{\basename}{boostToParavector}
%\renewcommand{\dirname}{notes/phy1520/}
\renewcommand{\dirname}{gabookI/}
%\newcommand{\dateintitle}{}
%\newcommand{\keywords}{}

\input{../latex/peeter_prologue_print2.tex}

\usepackage{peeters_layout_exercise}
\usepackage{peeters_braket}
\usepackage{peeters_figures}
\usepackage{siunitx}
%\usepackage{mhchem} % \ce{}
%\usepackage{macros_bm} % \bcM
%\usepackage{macros_qed} % \qedmarker
%\usepackage{txfonts} % \ointclockwise

\beginArtNoToc

\generatetitle{Lorentz boosts in GA paravector notation.}
%\chapter{Boosts in paravector notation}
%\label{chap:boostToParavector}
% \citep{sakurai2014modern} pr X.Y
% \citep{pozar2009microwave}
% \citep{qftLectureNotes}
% \citep{doran2003gap}
% \citep{jackson1975cew}
% \citep{griffiths1999introduction}

\section{Motivation.}

The notation I prefer for relativistic geometric algebra uses Hestenes' space time algebra (STA) \citep{doran2003gap}, where the basis is a four dimensional space \( \setlr{ \gamma_\mu } \), subject to Dirac matric like relations \( \gamma_\mu \cdot \gamma_\nu = \eta_{\mu \nu} \).

In this formalism a four vector is just the sum of the products of coordinates and basis vectors, for example, using summation convention

\begin{dmath}\label{eqn:boostToParavector:160}
x = x^\mu \gamma_mu.
\end{dmath}

The invarient for a four-vector in STA is just the square of that vector

\begin{dmath}\label{eqn:boostToParavector:180}
x^2
= (x^\mu \gamma_\mu) \cdot (x^\nu \gamma_\nu)
= \sum_\mu (x^\mu)^2 (\gamma_\mu)^2
= (x^0)^2 - \sum_{k = 1}^3 (x^k)^2
= (ct)^2 - \Bx^2.
\end{dmath}

Recall that a four-vector is time-like if this squared-length is positive, spacelike if negative, and light-like when zero.

Time-like projections are possible by dotting with the ``lab-frame'' time like basis vector \( \gamma_0 \)

\begin{equation}\label{eqn:boostToParavector:200}
ct = x \cdot \gamma_0 = x^0,
\end{equation}

and space-like projections are wedges with the same

\begin{equation}\label{eqn:boostToParavector:220}
\Bx = x \cdot \gamma_0 = x^k \sigma_k,
\end{equation}

where sums over latin indexes \( k \in \setlr{1,2,3} \) are implied, and where the elements \( \sigma_k \)

\begin{dmath}\label{eqn:boostToParavector:80}
\sigma_k = \gamma_k \gamma_0.
\end{dmath}

which are bivectors in STA,
can be viewed as an Euclidean vector basis \( \setlr{ \sigma_k } \).

Rotations in STA involve exponentials of space like bivectors \( \theta = a_{ij} \gamma_i \wedge \gamma_j \)

\begin{equation}\label{eqn:boostToParavector:240}
x' = e^{ \theta/2 } x e^{ -\theta/2 }.
\end{equation}

Boosts, on the other hand, have exactly the same form, but the exponenentials are with repsect to space-time bivectors arguments, such as \( \theta = a \wedge \gamma_0 \), where \( a \) is any four-vector.

Observe that both boosts and rotations neccessarily conserve the space-time length of a four vector (or any multivector with a scalar square).

\begin{dmath}\label{eqn:boostToParavector:260}
\lr{x'}^2
=
\lr{ e^{ \theta/2 } x e^{ -\theta/2 } } \lr{ e^{ \theta/2 } x e^{ -\theta/2 } }
=
e^{ \theta/2 } x \lr{ e^{ -\theta/2 } e^{ \theta/2 } } x e^{ -\theta/2 }
=
e^{ \theta/2 } x^2 e^{ -\theta/2 }
=
x^2 e^{ \theta/2 } e^{ -\theta/2 }
=
x^2.
\end{dmath}

\section{Paravectors.}

Paravectors, as used by Baylis \citep{baylis2004electrodynamics}, represent four-vectors using a Euclidean multivector basis \( \setlr{ \Be_\mu } \), where \( \Be_0 = 1 \).  The conversion between STA and paravector notation requires only multiplication with the timelike basis vector for the lab frame \( \gamma_0 \)

\begin{dmath}\label{eqn:boostToParavector:40}
X
=
x \gamma_0
= \lr{ x^0 \gamma_0 + x^k \gamma_k } \gamma_0
= x^0 + x^k \gamma_k \gamma_0
= x^0 + \Bx
= c t + \Bx,
\end{dmath}

We need a different structure for the invariant length in paravector form.  That invariant length is
\begin{dmath}\label{eqn:boostToParavector:280}
x^2
=
\lr{ \lr{ ct + \Bx } \gamma_0 }
\lr{ \lr{ ct + \Bx } \gamma_0 }
=
\lr{ \lr{ ct + \Bx } \gamma_0 }
\lr{ \gamma_0 \lr{ ct - \Bx } }
=
\lr{ ct + \Bx }
\lr{ ct - \Bx }.
\end{dmath}

Baylis introduces an involution operator \( \overbar{M} \) which toggles the sign of any vector or bivector grades of a multivector.  So if \( M = a + \Ba + I \Bb + I c \), is a multivector with all grades \( 0,1,2,3 \), then the involution of \( M \) is

\begin{dmath}\label{eqn:boostToParavector:300}
\overbar{M} = a - \Ba - I \Bb + I c.
\end{dmath}

Utilizing this operator, the invarient length for a paravector \( X \) is \( X \overbar{X} \).

Let's consider how boosts and rotations can be expressed in the paravector form.  The half angle operator for a boost along the spacelike \( \Bv = v \vcap \) direction has the form

\begin{dmath}\label{eqn:boostToParavector:120}
L = e^{ -\vcap \phi/2 },
\end{dmath}

\begin{dmath}\label{eqn:boostToParavector:140}
X'
=
c t' + \Bx'
=
x' \gamma_0
=
L x L^\dagger
=
e^{ -\vcap \phi/2 } x^\mu \gamma_\mu
e^{ \vcap \phi/2 } \gamma_0
=
e^{ -\vcap \phi/2 } x^\mu \gamma_\mu \gamma_0
e^{ -\vcap \phi/2 }
=
e^{ -\vcap \phi/2 } \lr{ x^0 + \Bx } e^{ -\vcap \phi/2 }
=
L X L.
\end{dmath}

Because the involution operator toggles the sign of vector grades, it is easy to see that the required invariance is maintained

\begin{dmath}\label{eqn:boostToParavector:320}
X' \overbar{X'}
=
L X L
\overbar{ L X L }
=
L X L
\overbar{ L } \overbar{ X } \overbar{ L }
=
L X \overbar{ X } \overbar{ L }
=
X \overbar{ X } L \overbar{ L }
=
X \overbar{ X }.
\end{dmath}

Let's explicitly expand the transformation of \cref{eqn:boostToParavector:140}, so we can relate the rapidity angle \( \phi \) to the magnitude of the velocity.

%}
\EndArticle
%\EndNoBibArticle
